---
title: "Safe data administration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Safe data administration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(entropiaR)
```

# v1 is read-only, by design

The EntropIA SQLite database is not a plain table store. It is guarded by a
sync engine and **81 triggers** (48 oplog-capture triggers maintain
`sync_row_versions`, and 33 collection-activity triggers maintain
`collections.updated_at`), and it may be *live* under the EntropIA desktop app
at the moment you connect. A bug in a write path could corrupt a database that
took years to build.

So v1 makes writes impossible rather than merely discouraged:

```{r}
con <- entropia_connect(system.file("extdata", "entropia-example.sqlite", package = "entropiaR"))
```

- the connection is opened with SQLite's read-only flags and
  `PRAGMA query_only = ON` is re-asserted on open;
- the write API exists as **stubs** with the v2 signatures, but every one
  errors with `entropia_error_write_disabled` before touching the database;
- `entropia_connect(write = TRUE)` is rejected at the door.

```{r error = TRUE}
entropia_insert(con, "items", data.frame(title = "nope"))
```

```{r error = TRUE}
entropia_connect(system.file("extdata", "entropia-example.sqlite", package = "entropiaR"),
                 write = TRUE)
```

Every stub error message names the verb, states that v1 is read-only, points at
the v2 plan, and suggests `entropia_connect(path, write = TRUE)` for v2.

# WAL and the live database

While EntropIA runs, the database may have `-wal` and `-shm` sidecar files and
be journaling in WAL mode. Reading through WAL is safe — a read-only client sees
a consistent snapshot. Two practical rules:

- **Never** copy the database with a file copy while sidecars exist; use
  `entropia_copy()`, which reads *through* the sidecars and emits one
  self-contained snapshot via `VACUUM INTO`.
- If the database is busy, `entropia_connect()` raises
  `entropia_error_locked` with guidance to retry or snapshot.

```{r}
copy_path <- tempfile(fileext = ".sqlite")
entropia_copy(con, copy_path)
```

# Schema compatibility policy

Because the database evolves, the package checks the schema on open and acts on
`options(entropiaR.schema_policy)`:

- `"warn"` (default) — warn and proceed for unknown/newer/older schemas,
  as long as required columns exist;
- `"error"` — hard stop on an unknown (newer) schema;
- `"allow"` — silent.

```{r}
entropia_schema_compat(con)$status
```

# The v2 write design

Full write support is designed and documented here; it is implemented in v2.
The design is deliberately conservative:

**Connection.** `entropia_connect(path, write = TRUE)` will open read-write,
set `PRAGMA foreign_keys = ON`, and warn loudly when the database may be live
in EntropIA.

**Verbs — all transactional, all guarded:**

```r
entropia_insert(con, table, data, dry_run = TRUE)
entropia_update(con, table, data, by, dry_run = TRUE)
entropia_upsert(con, table, data, by, dry_run = TRUE)
entropia_delete(con, table, filter, all = FALSE, confirm = FALSE)
```

- `dry_run = TRUE` is the default for writes: validate against the column
  contract, show what would change, change nothing.
- `entropia_update()` requires `by` and never touches primary keys.
- `entropia_upsert()` is implemented as `INSERT ... ON CONFLICT(id) DO UPDATE`
  — **never** `INSERT OR REPLACE`, because the sync engine bans it (rowid
  reassignment breaks FTS5 rowid joins).
- `entropia_delete()` requires a `filter`; deleting a whole table needs
  `all = TRUE` *and* `confirm = TRUE`.
- Every verb runs inside `DBI::dbWithTransaction()`.
- Statements are parameterized only; `INSERT OR REPLACE` and manual
  `PRAGMA` fiddling are outside the supported surface.
- A global guard, `options(entropiaR.write_dry_run = TRUE)`, forces dry-run for
  the whole session.

**Triggers are respected, never fought.** The 81 triggers are part of the
schema contract. `collections.updated_at`, for example, is trigger-maintained —
a v2 writer would not touch it directly. Validation runs against the same
column contract the read path uses, so a write can never introduce a row the
reader cannot understand.

# What this means for you today

For v1 you can treat the database as immutable input. Administer it with the
EntropIA app, snapshot it with `entropia_copy()` for analysis, and know that
`entropiaR` cannot corrupt it. When v2 lands, the same verbs will be the safe,
validated path — but until then the stubs are the guarantee.

```{r}
entropia_disconnect(con)
```
