---
title: "A complete reproducible analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A complete reproducible analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
library(entropiaR)
library(dplyr)
```

This vignette walks one end-to-end analysis: snapshot the database, build a
provenance-stamped dataset, profile it, and export — everything documented so
the result can be reproduced later.

# 1. Snapshot, then connect

The database may be live under EntropIA. Before analysis, take a snapshot with
`entropia_copy()` and run against *that*, so the analysis is immune to writes
landing mid-run and the provenance records a stable source.

```{r}
con <- entropia_connect(system.file("extdata", "entropia-example.sqlite", package = "entropiaR"))
snap <- tempfile(fileext = ".sqlite")
entropia_copy(con, snap)
snap_con <- entropia_connect(snap)
entropia_disconnect(con) # original connection no longer needed
```

# 2. Build the analysis dataset

`entropia_analysis_dataset()` applies filters to the corpus, collects it, and
stamps provenance. This is the boundary: everything after it is ordinary
tibble work, and everything before it is recorded.

```{r}
corpus <- entropia_analysis_dataset(snap_con, name = "corpus_full")
corpus
```

```{r}
entropia_provenance(corpus)
```

# 3. Temporal profile

When were assets created? `entropia_temporal_profile()` buckets a date column.
Collect the asset accessor directly — `entropia_collect()` types its
millisecond timestamps to `POSIXct`, which the profiler requires:

```{r}
assets <- entropia_assets(snap_con) |> entropia_collect()
temporal <- entropia_temporal_profile(assets, created_at, unit = "second")
temporal
```

# 4. Document lengths

`entropia_document_lengths()` appends character and word counts per document to
any tibble carrying a text column:

```{r}
texts <- entropia_text(snap_con) |> entropia_collect()
lengths <- entropia_document_lengths(texts)
lengths |> select(id, n_chars, n_words)
```

Empty text counts as 0; `NA` text stays `NA`.

# 5. Entities

`entropia_entities()` excludes soft-deleted rows by default and surfaces
provenance (source model) per entity. `entropia_entity_frequency()` tallies
them by type:

```{r}
entities <- entropia_entities(snap_con) |> entropia_collect()
entropia_entity_frequency(entities)
```

# 6. Topics

Topics are normalized UPPERCASE names. Join `item_topics` to `topics`, then
tally:

```{r}
item_topics <- entropia_item_topics(snap_con) |> entropia_collect()
topics <- entropia_topics(snap_con) |> entropia_collect()
entropia_topic_frequency(left_join(item_topics, topics, by = c("topic_id" = "id")))
```

# 7. Collections side by side

`entropia_compare_collections()` summarizes the corpus per collection:

```{r}
entropia_compare_collections(corpus)
```

# 8. Corpus quality

`entropia_corpus_quality()` reports OCR coverage, transcription presence,
and empty texts per asset type, plus metadata coverage per collection:

```{r}
quality <- entropia_corpus_quality(snap_con)
quality
```

# 9. Visualize

`entropia_plot_*` helpers wrap the analysis summaries in ggplot2 (Suggests) and
return ordinary `ggplot` objects you can extend:

```{r}
entropia_plot_temporal(temporal)
```

```{r}
entropia_plot_entities(entropia_entity_frequency(entities))
```

```{r}
entropia_plot_coverage(quality)
```

# 10. Export with provenance

The dataset exports as-is (RDS keeps its class and provenance; CSV is streamed
for lazy inputs). Always write the provenance sidecar alongside:

```{r}
export_path <- tempfile(fileext = ".csv")
entropia_export(entropia_corpus(snap_con), export_path)

rds_path <- tempfile(fileext = ".rds")
entropia_export(corpus, rds_path, format = "rds")
prov_path <- tempfile(fileext = "-prov.json")
entropia_write_provenance(corpus, prov_path)
```

The RDS round-trips with its provenance intact:

```{r}
back <- readRDS(rds_path)
entropia_provenance(back)[["name"]]
```

# Cleanup

```{r}
entropia_disconnect(snap_con)
```

That is the loop: snapshot → build a stamped dataset → analyze → export with a
sidecar. Each step uses functions documented elsewhere in this site; the
provenance records the whole chain.
