---
title: "Exploring the corpus"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring the corpus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(entropiaR)
library(dplyr)
```

# The domain model

EntropIA organizes your research material in a three-level hierarchy:

```
collections ──→ items ──→ assets
```

- a **collection** is a research project or folder (e.g. "Conflicto SOIP 1965-66");
- an **item** is one document inside a collection (a newspaper page, a letter, a photo);
- an **asset** is one physical file belonging to an item — an image, a PDF, or an
  audio recording. A PDF item usually has one asset *per page*; PDF page assets
  point at their parent asset via `parent_asset_id` and carry a `page_number`.

Everything else hangs off this spine: text layers (`extractions`,
`transcriptions`, `layouts`), entities and triples, notes and annotations,
topics, LLM results, and RAG artifacts.

# Accessors are lazy

Every accessor returns a lazy `tbl_sql` — a query that has *not* run yet. SQLite
does the work when you collect. Nothing is loaded into R memory until you ask:

```{r}
con <- entropia_connect(system.file("extdata", "entropia-example.sqlite", package = "entropiaR"))
items <- entropia_items(con)
items
```

Because these are `tbl_sql`, you can pipe them straight into dplyr verbs and the
whole query is translated to SQL (see `vignette("dplyr")`).

# The three spine accessors

```{r}
entropia_collections(con) |> collect()
```

```{r}
entropia_items(con) |> collect()
```

```{r}
entropia_assets(con) |> collect()
```

Notice the PDF pages: assets `...3332` and `...3333` carry a
`parent_asset_id` and a `page_number`, while the top-level PDF asset does not.
You can join pages back to their parent:

```{r}
assets <- entropia_assets(con) |> collect()
assets |>
  filter(!is.na(page_number)) |>
  select(asset_id = id, page_number, parent_asset_id) |>
  left_join(select(assets, id, type), by = c("parent_asset_id" = "id"))
```

The asset accessor never selects BLOB columns such as `embedding` — those are
opt-in (see `entropia_embeddings()` / `entropia_chunks()`).

# The corpus view

`entropia_corpus()` is the workhorse: it joins items, collections and assets in
one lazy query, one row per asset, and appends the best available text per
asset:

```{r}
corpus <- entropia_corpus(con)
corpus |> colnames()
```

```{r}
corpus |>
  select(item_title, asset_type, page_number, text) |>
  collect()
```

Assets with no text layer keep their row and get `NA` text — no rows are lost.

## Filtering

`collections` and `asset_types` accept character vectors and push the filter
down to SQL:

```{r}
entropia_corpus(con, asset_types = "pdf") |>
  collect() |>
  nrow()
entropia_corpus(con, collections = "Archivo de prueba") |>
  collect() |>
  nrow()
```

By default the corpus includes PDF *page* assets. Set `page_assets = FALSE` to
see only top-level assets:

```{r}
entropia_corpus(con) |>
  collect() |>
  nrow()
entropia_corpus(con, page_assets = FALSE) |>
  collect() |>
  nrow()
```

`include_deleted` is reserved for the v2 read-write release; in v1 the flag has
no effect (the corpus carries no soft-delete marker).

# Item metadata

`entropia_metadata()` parses the JSON inside `items.metadata` into tidy rows:
the `__entropia_file_metadata` fields become proper columns (`original_name`,
`original_path`, `imported_at` as a `POSIXct`), and any remaining top-level keys
become list-columns. Items without metadata get a row with `NA`s rather than
disappearing:

```{r}
entropia_metadata(con)
```

Pass `parse = FALSE` to see the raw JSON text instead.

# Cleaning up

```{r}
entropia_disconnect(con)
```

Next: `vignette("text")` for text extraction and metadata, or
`vignette("dplyr")` for composing lazy queries.
