---
title: "Reproducible analysis datasets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducible analysis datasets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(entropiaR)
```

# Why provenance?

A research corpus changes over time: EntropIA keeps syncing, the schema
evolves, and a "just filter and collect" workflow makes it impossible to say
later *which* data produced a result. `entropiaR` solves this at the boundary
where a query becomes a dataset: `entropia_analysis_dataset()` records exactly
what went in.

```{r}
con <- entropia_connect(system.file("extdata", "entropia-example.sqlite", package = "entropiaR"))
```

# Building a dataset

`entropia_analysis_dataset()` takes a connection, applies filter expressions to
the corpus, collects it, and stamps provenance:

```{r}
ds <- entropia_analysis_dataset(con, name = "pdf_documents", asset_type == "pdf")
ds
```

The result is a tibble with class `entropia_dataset`, plus an `entropia_prov`
attribute recording the schema version, the content hash of the source schema,
the source path, the filter expressions, the package version, a build timestamp,
and the R version.

```{r}
entropia_provenance(ds)
```

Multiple filters compose, and they push down to SQL before the collect:

```{r}
entropia_analysis_dataset(con, name = "audio_items", asset_type == "audio") |>
  nrow()
```

# Determinism

Identical inputs produce identical datasets. The corpus is always arranged on
`asset_id` before collecting, so row order never depends on the physical layout
of the database:

```{r}
a <- entropia_analysis_dataset(con, asset_type == "image")
b <- entropia_analysis_dataset(con, asset_type == "image")
identical(a, b) # data bytes identical
identical(
  entropia_provenance(a)[["content_hash"]],
  entropia_provenance(b)[["content_hash"]]
)
```

Only `built_at` differs between builds.

# Provenance sidecars

`entropia_write_provenance()` writes the provenance as a JSON sidecar, so a
dataset and its `-prov.json` file travel together and survive email, archives,
and other people's machines:

```{r}
prov_path <- tempfile(fileext = "-prov.json")
entropia_write_provenance(ds, prov_path)
readLines(prov_path)[1:9]
```

# Exporting

`entropia_export()` writes a dataset or a lazy query to disk. Delimited formats
(csv/tsv) are *streamed* for lazy inputs — they fetch in chunks and never
collect the whole result — and the output is deterministically ordered. JSON,
RDS, parquet and arrow are whole-file formats.

```{r}
csv_path <- tempfile(fileext = ".csv")
entropia_export(entropia_corpus(con), csv_path) # lazy input, streamed
read.csv(csv_path) |> nrow()
```

```{r}
rds_path <- tempfile(fileext = ".rds")
entropia_export(ds, rds_path, format = "rds") # class + provenance intact
back <- readRDS(rds_path)
class(back)
```

```{r}
json_path <- tempfile(fileext = ".json")
entropia_export(entropia_collect(entropia_corpus(con)), json_path, format = "json")
jsonlite::fromJSON(json_path) |> nrow()
```

parquet/arrow are available when the `arrow` package is installed (it is a
Suggests dependency, so `entropia_export` fails with clear guidance rather than
a load error when it is not).

# Recommended workflow

For anything that will take a while, snapshot the database first, then build
datasets against the snapshot. The provenance then records the *snapshot* as
the source, so the analysis is immune to writes landing in the meantime:

```{r}
snap <- tempfile(fileext = ".sqlite")
entropia_copy(con, snap)
snap_con <- entropia_connect(snap)
stable <- entropia_analysis_dataset(snap_con, name = "stable", asset_type == "pdf")
entropia_provenance(stable)[["source_path"]] == snap
entropia_disconnect(snap_con)
```

# Cleaning up

```{r}
entropia_disconnect(con)
```

Next: `vignette("analysis")` for a complete end-to-end analysis, or
`vignette("administration")` for the safety model around the database.
