---
title: "Lazy queries with dplyr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Lazy queries with dplyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(entropiaR)
library(dplyr)
```

# Everything is a lazy table

Every accessor returns a `tbl_sql` backed by SQLite. That means you can compose
the whole `dplyr` grammar — `filter`, `select`, `mutate`, `arrange`,
`group_by`, `summarise`, `left_join` — and the *entire query* is translated to
SQL and executed by SQLite. Nothing is loaded into R until you `collect()`.

```{r}
con <- entropia_connect(system.file("extdata", "entropia-example.sqlite", package = "entropiaR"))

entropia_items(con) |>
  filter(collection_id == "11111111-1111-4111-8111-111111111111") |>
  select(id, title) |>
  collect()
```

The `entropiaR` package deliberately defines *no* custom verbs. `dplyr` works
because the accessors are plain `tbl_sql` — zero magic, and you can read dbplyr
documentation for the full verb vocabulary.

# Filters and selection push down

This never pulls the `items` table into memory:

```{r}
q <- entropia_items(con) |>
  filter(grepl("huelga", title, ignore.case = TRUE)) |>
  select(id, title, created_at)
```

You can inspect the generated SQL to confirm the work stays on SQLite:

```{r}
dbplyr::sql_render(q)
```

tidyselect works natively too — the `select()` above could have been written
with helpers like `starts_with()`:

```{r}
entropia_items(con) |> select(id, starts_with("created")) |> collect()
```

# Joins across accessors

Because every accessor is a `tbl_sql`, you can join them directly. Entities
join to items for context:

```{r}
entropia_entities(con) |>
  left_join(entropia_items(con), by = c("item_id" = "id")) |>
  select(entity_type, value, title) |>
  collect()
```

Note that `entropia_entities()` excludes soft-deleted rows by default
(`source = 'manual_deleted'`), and `min_confidence` filters on confidence — both
push down to SQL.

# Grouped summaries

```{r}
entropia_assets(con) |>
  group_by(type) |>
  summarise(n = n(), total_size = sum(size, na.rm = TRUE)) |>
  arrange(desc(n)) |>
  collect()
```

# Parameter-safe full-text search

`entropia_search()` wraps SQLite's FTS5 index and is injection-safe: your query
text is escaped via `dbQuoteString` before it is spliced into `MATCH`. Results
come back ranked by BM25.

```{r}
entropia_search(con, "huelga") |> collect()
```

`index = "chunks"` searches the RAG chunk index instead:

```{r}
entropia_search(con, "huelga", index = "chunks") |> collect()
```

Multi-word queries are AND-ed, `limit` truncates the ranked result, and the
query itself is never interpolated into SQL.

# Collecting applies the column contract

`entropia_collect()` is `collect()` plus the package's column contract:
millisecond timestamps become `POSIXct`, JSON-in-TEXT columns become
list-columns, and embedding BLOBs stay `raw` (and are not selected unless you
opt in).

```{r}
items <- entropia_items(con) |> entropia_collect()
items$created_at
```

```{r}
metadata_col <- entropia_items(con) |> entropia_collect()
metadata_col$metadata
```

The pure timestamp helpers back this: `entropia_datetime()` converts
milliseconds, `entropia_datetime_s()` seconds, and `entropia_datetime_auto()`
uses a magnitude guard for columns whose units have drifted over time (used for
`entities.created_at` and `triples.created_at`):

```{r}
entropia_datetime(1768478400000)
entropia_datetime_s(1768478400)
entropia_datetime_auto(c(1768478400000, 1768478400))
```

# Cleaning up

```{r}
entropia_disconnect(con)
```

Next: `vignette("datasets")` for building reproducible analysis datasets, or
`vignette("analysis")` for a complete analysis workflow.
