---
title: "Exploratory analysis of an EntropIA corpus"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploratory analysis of an EntropIA corpus (English)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
library(entropiaR)
library(dplyr)
```

*English version.* Spanish: `vignette("eda")`.

Start with a snapshot, then one shared overview. Do not collect the full
corpus until a table of counts tells you the universe is the one you want.

```{r}
con <- entropia_connect(system.file(
  "extdata", "entropia-example.sqlite",
  package = "entropiaR"
))
```

# Inventory without loading text

`entropia_overview()` aggregates in SQLite. It never materialises OCR text,
JSON metadata or embedding BLOBs.

```{r}
eda <- entropia_overview(con)
names(eda)
eda$counts
eda$collections
eda$asset_types
```

`counts` uses explicit units (`item`, `asset`, `collection`). `n` in
`collections` is the number of *rows in the study query* (assets plus items
without an asset), not a synonym for `n_items`.

# One universe, many tables

Every table in `eda` uses the same selection. Filter by asset type or drop
PDF page assets without rewriting joins:

```{r}
pdfs <- entropia_overview(con, asset_types = "pdf", page_assets = FALSE)
pdfs$counts
pdfs$selection
```

An empty ID vector selects nothing (it does not fall back to “all”):

```{r}
entropia_overview(con, collection_ids = character())$counts
```

# Quality: numerators, denominators, status

```{r}
eda$quality
```

Read `status` before `pct`:

- `ok` — the rate is defined;
- `empty` — eligible rows exist but the field/layer is empty;
- `no_data` — no rows in the denominator;
- `not_applicable` — the metric does not apply to that group (e.g. OCR on
  audio, depending on eligibility);
- `invalid` — JSON that does not parse.

`metadata_coverage` is *presence* of metadata text. `metadata_validity` is
JSON syntax, not “the document has an author and a date”.

# Entities and topics: occurrence vs prevalence

```{r}
eda$entities
eda$topics
```

`n` is the number of rows (mentions). `n_items` is distinct items. `pct` is
`n_items / total` where `total` is the number of items in the universe —
items with no entity still sit in the denominator.

Soft-deleted entities (`source = "manual_deleted"`) stay out unless you pass
that source explicitly.

# Temporal profile and exclusions

```{r}
eda$temporal
attr(eda$temporal, "exclusions")
```

Buckets are UTC month starts of the chosen `date_var` (default
`item_created_at`). That column is an *operational* timestamp (created /
imported), not necessarily the date of the historical document.

# Profile a collected tibble

Overview stays on SQL. Local distributions need a collected table:

```{r}
lengths <- entropia_text(con) |>
  entropia_collect() |>
  entropia_document_lengths()
entropia_profile(lengths, columns = c("n_chars", "n_words"))
```

`entropia_profile()` returns `structure`, `missing`, `numeric`,
`categorical`, `duplicates` and `sampling`. Numeric ID columns are described
but not treated as scientific measurements in correlation plots (see
`vignette("visualize")`).

Sampling is opt-in and does not change `.Random.seed`:

```{r}
set.seed(1)
before <- .Random.seed
entropia_profile(lengths, sample_n = 2, seed = 99)$sampling$n
identical(.Random.seed, before)
```

# Hand-off

The same `eda` object feeds ggplot helpers, the Shiny dashboard and
`entropia_report()` — do not recompute a different filter per panel.

```{r}
entropia_disconnect(con)
```

Next: `vignette("visualize")`, or `vignette("analysis")` for a full
snapshot → dataset → export loop.
