---
title: "Extracting texts and metadata"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extracting texts and metadata (English)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(entropiaR)
```

*English version.* Spanish: `vignette("text")`.

# The text layers

An asset can have up to three machine-readable text layers, stored in three
1:1 tables:

| Table | Meaning | Main columns |
| --- | --- | --- |
| `extractions` | OCR / PDF text extraction | `text_content`, `method`, `confidence` |
| `transcriptions` | Speech-to-text for audio | `text_content`, `language`, `duration_ms`, `segments` |
| `layouts` | Layout analysis (regions/blocks) | `regions`, `blocks`, `image_width`, `image_height` |

Each row is keyed deterministically to its asset (`ext-`, `trx-`, `lay-` plus
the asset id), so the layer tables join 1:1 to `assets` with no fan-out.

```{r}
con <- entropia_connect(system.file("extdata", "entropia-example.sqlite", package = "entropiaR"))
entropia_extractions(con) |> entropia_collect()
entropia_transcriptions(con) |> entropia_collect()
entropia_layouts(con) |> entropia_collect()
```

# The best text per asset

`entropia_text()` returns one row per asset with a single `text` column — the
best available text per asset. `source` selects the layer:

- `"extraction"` — the OCR/PDF extraction text only;
- `"transcription"` — the audio transcription text only;
- `"auto"` (default) — extraction **when that layer exists**, even if the
  extraction text is empty, otherwise transcription. This matches the
  corpus `COALESCE` rule. Coverage metrics that mean “any layer useful”
  are named separately in `entropia_overview()$quality`.

```{r}
entropia_text(con) |> entropia_collect()
```

Assets with neither layer keep their row and get `NA` text:

```{r}
entropia_text(con) |>
  entropia_collect() |>
  dplyr::filter(is.na(text))
```

# OCR markers

EntropIA's PDF extraction embeds *image markers* in the text —
`![](page=n,bbox=[...])` — one per embedded page image. They are noise for text
analysis, so `entropia_text()` strips them by default. See the raw text first:

```{r}
raw <- entropia_text(con, strip_markers = FALSE) |> entropia_collect()
raw$text[1]
```

And the stripped version:

```{r}
clean <- entropia_text(con) |> entropia_collect()
clean$text[1]
```

The stripping happens in SQL (a recursive CTE), so it stays lazy and
composable — you can `filter`/`mutate` on top of `entropia_text()` and the whole
query still runs on SQLite.

# Restricting to specific assets

`assets` accepts a character vector of asset ids or a lazy `tbl_sql`:

```{r}
ids <- entropia_assets(con) |>
  dplyr::filter(type == "audio") |>
  dplyr::pull(id)
entropia_text(con, assets = ids) |> entropia_collect()
```

# JSON columns become list-columns

`entropia_collect()` applies the column contract. JSON-in-TEXT columns become
list-columns via `jsonlite`, so `transcriptions.segments` (an array of
`{start_ms, end_ms, text}` objects) comes back as a list of data frames:

```{r}
trx <- entropia_transcriptions(con) |> entropia_collect()
trx$segments[[1]]
```

The same applies to `items.metadata`, `rag_messages.sources`, and
`llm_results.result`.

# Metadata

`entropia_metadata()` parses `items.metadata` into tidy rows (see
`vignette("corpus")`). Non-scalar file-metadata fields no longer abort the
whole batch: the row is kept, the original JSON stays in `raw_metadata`, and
`attr(..., "diagnostics")` lists `item_id`, field and problem.

```{r}
entropia_metadata(con)
```

Malformed JSON never fails the collect — it warns (with row context) and
yields `NA`. Use `parse = FALSE` for the raw text.

# Coverage at a glance

To see which assets are missing a text layer, use the coverage helpers:

```{r}
entropia_ocr_coverage(con, by = "asset") |> entropia_collect()
```

Or the combined quality report:

```{r}
entropia_corpus_quality(con)
```

# Cleaning up

```{r}
entropia_disconnect(con)
```

Next: `vignette("dplyr")` for composing lazy queries, or
`vignette("datasets")` for building reproducible analysis datasets.
