---
title: "Plotting EntropIA summaries"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting EntropIA summaries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 3.8)
library(entropiaR)
```

Plot helpers consume **already-summarised tibbles**. They never query SQLite.
ggplot2 is in Suggests: install it to plot; the rest of the package works
without it.

```{r}
con <- entropia_connect(system.file(
  "extdata", "entropia-example.sqlite",
  package = "entropiaR"
))
eda <- entropia_overview(con)
```

# Collections and coverage

```{r}
entropia_plot_collections(eda$collections)
entropia_plot_coverage(eda$quality, metric = "ocr_coverage")
```

Coverage uses `group` as the axis label and `group_id` when present, so two
collections named “Archivo” stay two bars. `status` of `no_data` /
`not_applicable` is annotated rather than drawn as 0%.

# Entities and topics

```{r}
entropia_plot_entities(eda$entities)
entropia_plot_topics(eda$topics)
```

`top` keeps the most frequent rows. `top_by = "group"` (entities only) takes
the top-N *inside each group* instead of globally.

# Time series

Overview temporal data uses a column named `date`:

```{r}
entropia_plot_temporal(eda$temporal, date_var = "date")
```

A profile built with `entropia_temporal_profile(..., by = collection_name)`
must pass `group` (or the helper errors when several leftover columns could
be the series):

```{r}
items <- entropia_collect(entropia_items(con))
by_title <- entropia_temporal_profile(items, created_at, unit = "day")
entropia_plot_temporal(by_title)
```

# Lengths and missingness

```{r}
lengths <- entropia_document_lengths(entropia_collect(entropia_text(con)))
entropia_plot_distribution(lengths, n_words, type = "histogram")
entropia_plot_distribution(lengths, n_chars, type = "boxplot")

prof <- entropia_profile(lengths, columns = c("n_chars", "n_words"))
entropia_plot_missing(prof$missing)
```

Empty input draws a “Sin datos…” annotation rather than crashing.

# Extend and save

Every helper returns a plain `ggplot`:

```{r}
p <- entropia_plot_entities(eda$entities) +
  ggplot2::labs(caption = "Ocurrencias, no identidades resueltas")
tmp <- tempfile(fileext = ".png")
ggplot2::ggsave(tmp, p, width = 6, height = 4)
file.exists(tmp)
```

```{r}
entropia_disconnect(con)
```

Next: `vignette("dashboard")` to put the same tables in a local app.
